---
title: "Cross-Sectional Analysis with a Single Wave"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-Sectional Analysis with a Single Wave}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## When to use this workflow

A cross-sectional design treats one wave as a population snapshot. This is
appropriate when you need a point-in-time estimate (prevalence, association)
and do not require within-person change. Common examples include a Master's
thesis using the most recent Health wave, or a methods paper demonstrating a
new estimator on a single data release.

Even for a single-wave analysis, lissr adds value: it automates
authentication, lets you pick exactly the files you need, and gives you
access to the merge engine's sentinel-code cleaning so you start from
analysis-ready data instead of raw SPSS.

## Step 1 — download one wave of one module

```{r download-single}
library(lissr)

# authenticate (once per session)
liss_login()

# build the file inventory
bp <- liss_blueprint()

# filter to the latest Health wave, SPSS format
latest_health <- bp |>
  dplyr::filter(
    module == "Health",
    type   == "spss"
  ) |>
  dplyr::filter(wave == max(wave))

latest_health
#> # A tibble: 1 × 8
#>   module module_id  wave wave_id type  name       file              path
#>   <chr>      <int> <int>   <int> <chr> <chr>      <chr>             <chr>
#> 1 Health        18    18    1102 spss  ch25r 1.0p ch25r_1_0p_EN.sav /down…

# download just that one file
liss_download(latest_health, .dir = "data/ch")
```

## Step 2 — read and clean the data

You can read the file directly with `haven`, or you can leverage the
merge engine to apply the recipe's sentinel-code recoding even for a
single wave. The recipe knows which numeric codes are "don't know" vs
"prefer not to say" vs genuine missing — cleaning you would otherwise
do by hand with a codebook open.

```{r clean-via-recipe}
library(haven)
library(dplyr)

# option A: read raw and clean yourself
raw <- haven::read_sav("data/ch/ch25r_1_0p_EN.sav")
dim(raw)
#> [1] 4892  271

# option B: use the merge engine for a single wave
# (this applies prefix stripping, sentinel recoding, and labelled policy)
recipe <- liss_recipe("ch")

# temporarily trim the recipe to just the wave you need
recipe$wave_index <- purrr::keep(
  recipe$wave_index,
  ~ .x$id == "ch25r"
)

result <- merge_liss_module(recipe, data_dir = "data/ch", output_dir = "output")
health <- result$data
dim(health)
#> [1] 4892  265
```

The merge engine's output has cleaner column names (no `ch24q` prefix),
sentinel values already recoded to `NA`, and a `wave_id` / `wave_year`
column appended.

## Step 3 — attach background variables

Most cross-sectional analyses need demographics: age, sex, education,
income. These live in the Background Variables file (`avars`, a
separate monthly release), not in the survey module itself.

**Critical rule**: join on `nomem_encr` only. Never use `nohouse_encr` —
household assignments change over time. Match the fieldwork month, not
the calendar year.

```{r attach-demographics}
# download the background variables file for the same fieldwork period
# (check fieldwork_ym in the merged output for the exact month)
bg_files <- bp |>
  dplyr::filter(
    module == "Background Variables",
    type   == "spss",
    wave   == 202511  # YYYYMM matching the fieldwork period
  )

liss_download(bg_files, .dir = "data/avars")

# each avars file also carries this period as its `wave` column
avars <- haven::read_sav("data/avars/avars_202511_EN_1_0p.sav") |>
  haven::zap_labels() |>
  dplyr::select(
    nomem_encr,
    age       = leeftijd,
    sex       = geslacht,
    edu_level = oplcat,
    hh_income = nettohh_f,
    urban     = sted
  )

analysis_df <- health |>
  dplyr::left_join(avars, by = "nomem_encr")

nrow(analysis_df)
#> [1] 4892
```

## Step 4 — run an analysis

With the merged, cleaned data you can proceed to standard modelling. For
example, estimating the association between education and self-rated
health (suffix `s001` in the harmonised output), adjusting for age and
sex:

```{r model}
analysis_df <- analysis_df |>
  dplyr::mutate(
    srh = factor(s001, levels = 1:5,
                 labels = c("poor", "moderate", "good", "very good", "excellent")),
    female = as.integer(sex == 2)
  )

fit <- MASS::polr(srh ~ edu_level + age + female, data = analysis_df)
summary(fit)
```

## When *not* to use this approach

If your research question involves change (did self-rated health improve
after a policy?), or if you want to exploit the panel structure for
causal identification (fixed effects, difference-in-differences), you
should move to the longitudinal workflow described in
[Longitudinal Panel Analysis](longitudinal-panel-analysis.html)
*(or run `vignette("longitudinal-panel-analysis", package = "lissr")` in the console)*.
A single wave cannot separate
age, period, and cohort effects, and it provides no within-person
variation to control for time-invariant confounders.

## Checklist

Before submitting results from a cross-sectional LISS analysis, verify:

- [ ] You joined background variables on `nomem_encr` (not `nohouse_encr`).
- [ ] The background variables file matches the fieldwork month, not the
  calendar year.
- [ ] Sentinel codes (-9 = don't know, -8 = prefer not to say, 999, etc.)
  have been recoded to `NA` — the merge engine does this automatically,
  but check if you loaded raw SPSS directly.
- [ ] You report the wave identifier (e.g. ch25r) and the LISS data
  version number in your methods section so results are reproducible.
- [ ] If you selected a single wave from a module with instrument changes
  (boundary rules), note that in your limitations.
