---
title: "Income Cleaning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Income Cleaning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Why clean income at all

Self-reported household income in the LISS panel contains a familiar
family of artefacts: misplaced decimal points, monthly amounts entered
in an annual field, personal income entered in the household field,
placeholder values, sign errors, and residual SPSS missing codes. Left
in place, a single misplaced zero moves a household across every income
bracket an analysis uses.

`liss_clean_income()` turns the cleaning procedures that used to live
in ad hoc analysis scripts into a rule-driven, fully audited stage of
the lissr pipeline. Every decision rule is declared in a YAML ruleset,
every applied decision lands in a ledger with its evidence and a
plain-language justification, the original values stay in the returned
data, and a generated report lets any reader reconstruct and contest
each modification.

## The ruleset

The default ruleset ships with the package and is the single source of
truth for what the cleaner may do:

```{r ruleset}
library(lissr)

rs <- liss_cleaning_ruleset()
print(rs)

system.file("cleaning", "income_cleaning_rules.yml", package = "lissr")
```

Rules are grouped into four sections. Preparation rules (P01 to P06)
resolve columns, attach background demographics, expand bracket codes,
rectify sign errors, and sweep residual SPSS missing codes. Detection
rules (D01 to D11) identify implausible cells, from unrecoverable
placeholder values through household-level scale errors and robust
statistical outliers to dataset-level extremes. Correction rules (C01
to C06) generate candidate replacement values. The finalization rule
(F01) handles anything still outside the plausible range according to
its `disposition`: `void` (the default) sets such values to NA,
`winsorise` clamps over-cap values to the cap, and `flag` only ledgers
them. Each rule carries an id, a description, a rationale, optional
literature references, parameters, and an `enabled` switch.

## Quick start: dry run first, then correct

The cleaner operates on merged module output. A dry run in `"flag"`
mode performs the full procedure without touching a single value; the
simulated result arrives in `nethh_proposed` and every would-be
decision in the ledger:

```{r flag-mode}
result <- merge_liss_module(liss_recipe("ci"), data_dir = "data/ci",
                            output_dir = "output/ci")

dry <- liss_clean_income(result, mode = "flag")

head(dry$decisions)
summary(dry)
```

Once the proposals look right, run in the default `"correct"` mode and
write the audit artifacts:

```{r correct-mode}
cleaned <- liss_clean_income(result, output_dir = "output/ci")

cleaned$data$nethh          # cleaned values
cleaned$data$nethh_observed # untouched input, always preserved
table(cleaned$data$nethh_clean_status, useNA = "ifany")
```

A third mode, `"na_only"`, voids detected cells without imputing
anything, for analyses that prefer missingness over model-based
corrections.

## Reading a decision

Every modified cell has at least one ledger row. A scale-error
correction looks like this:

```{r ledger}
d <- cleaned$decisions
d[d$rule_id == "D06", c("person_id", "wave", "observed", "corrected",
                        "candidate_source", "evidence")]

# the justification column carries the full sentence, for example:
# "detected by scale_error; replaced with the household_median candidate
#  25800 (closest of 4 admissible candidate(s) to the household_median
#  25800; constrained to [8000, 150000])"
```

The `candidates` column lists every admissible candidate the selection
saw, so a reader can verify that the applied value was the closest to
the anchor, and the `valid_min`/`valid_max` columns show the constraint
window (category bounds where reported, global limits otherwise).

## The report

`liss_cleaning_report()` renders three artifacts: a markdown report
whose methodology section is generated from the ruleset itself (every
rule with its description, rationale, parameters, and references), the
complete decision ledger as CSV, and an engine-shaped JSONL audit log:

```{r report}
liss_cleaning_report(cleaned, output_dir = "output/ci")
```

Because the methodology is generated from the ruleset that actually
ran, the report can never drift from the code.

## Disagreeing with a rule

Every rule can be switched off or re-parameterised per run, and the
overrides are recorded in the report:

```{r overrides}
# a stricter volatility requirement for scale-error detection,
# no extreme-z net, and a higher plausibility cap
cleaned2 <- liss_clean_income(
  result,
  income_cap = 175000,
  disable = c("D10"),
  params = list(D06 = list(volatility_min = 0.7))
)
```

For a lasting policy, copy the packaged YAML, edit it, and pass the
path; `validate_cleaning_ruleset()` enforces the schema:

```{r custom-ruleset}
cleaned3 <- liss_clean_income(result, ruleset = "my_income_rules.yml")
```

## Attaching background demographics

Donor-pool matching (correction rule C05) works best with household
size, position, occupation, age, and education attached. Rule P01
aligns a monthly background file to the annual wave scale, keeps one
row per person-year, and joins on the person id, never on the household
id:

```{r background}
background <- haven::read_sav("data/avars_201801_EN_1.0p.sav")
cleaned <- liss_clean_income(result, background = background)
```

## Equivalised income

For cross-household comparison, `liss_equivalise_income()` provides the
scale the source analysis pipelines used, alongside the OECD-modified
and square-root alternatives:

```{r equivalise}
library(magrittr)

panel <- cleaned$data %>%
  dplyr::mutate(
    stand_inc = liss_equivalise_income(nethh, aantalhh, aantalki)
  )
```

One definitional caveat for the OECD-modified scale: it defines
children as household members under 14, while `aantalki` counts
children living at home of any age, so passing `aantalki` approximates
that scale; see `?liss_equivalise_income`.

## Where the rules came from

The ruleset consolidates the income-cleaning procedures of two
production analysis projects and corrects several latent defects found
while porting them; `INCOME_CLEANING_DESIGN.md` in the repository maps
every legacy construct to its rule, documents each deviation, and
describes how to propose new rules or parameter changes through pull
requests, so the methodology can improve under community review while
the audit trail keeps every historical run reproducible.
