---
title: "Merging LISS Panel Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Merging LISS Panel Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

lissr includes a recipe-driven merge engine that processes YAML
specifications conforming to the Canonical Schema v1.1.0 (a strictly
additive extension of v1.0.0; recipes may declare either). Each recipe
encodes all merge-relevant decisions for a module: file patterns,
variable harmonization, boundary handling, comparability contracts, and
validation checks.

All merged output is written in SPSS `.sav` format to preserve variable
labels, value labels, and user-defined missing values.

## Single module merge

```{r}
library(lissr)

recipe <- liss_recipe("ch")
result <- merge_liss_module(
  recipe,
  data_dir   = "liss/ch",
  output_dir = "./output"
)
```

This produces four files:

- `ch_merged.sav` — merged data (SPSS format, preserving all labels)
- `ch_merge_log.jsonl` — audit-grade structured log
- `ch_merge_summary.json` — per-run summary (if enabled in recipe)
- `ch_merge_report.txt` — human-readable report

The returned object additionally carries `provenance` (package, recipe,
and schema versions, per-input md5 hashes, release-selection decisions)
and a `valid_for_analysis` verdict, both reproduced in the report.

## Batch merge

```{r}
modules <- c("ch", "cv", "cd", "cf", "cw", "cp", "cs", "ci", "ca", "cr")
recipe_paths <- purrr::map_chr(modules, ~ {
  system.file("recipes", paste0(.x, "_merge_recipe.yml"), package = "lissr")
})

results <- merge_liss_modules(
  recipe_paths,
  data_dir   = "liss",
  output_dir = "./output"
)
```

## Cross-module panel merge

After merging individual modules, combine them into one wide dataset.
Columns are prefixed with the module code to avoid collisions
(e.g. `ch_s004`, `cv_s004`).

```{r}
panel <- merge_liss_panel(results, write_to = "./output/liss_panel.sav")

# only respondent-years present in all modules
panel_inner <- merge_liss_panel(results, join_type = "inner")
```

## Validate recipes

```{r}
recipe <- liss_recipe("ch")
validate_recipe(recipe, "ch_merge_recipe.yml")
```

## Onboard a new wave

When a new wave is released, the onboarding helper automates most of the
checklist: variable diffs, candidate wave_index entry, expected-presence
checks, and boundary alerts.

```{r}
onboard_new_wave(
  recipe_path  = system.file("recipes", "ch_merge_recipe.yml", package = "lissr"),
  new_file     = "ch25r_EN_1.0p.sav",
  prev_wave_id = "ch24q"
)
```

## Background variables

The Background Variables file (`avars`) is a monthly snapshot used as
the linkage backbone for all other modules. It is a separate release,
not one of the ten core modules. When merging background variables with
survey data:

1. Use only `nomem_encr` as the join key (never `nohouse_encr`).
2. Match the fieldwork month of the survey, not the calendar year.

The merged output includes a `fieldwork_ym` column (YYYYMM integer)
derived automatically from the LISS `_m` suffix. Each `avars` file
carries its fieldwork period as a `wave` column (YYYYMM), so stacking
monthly files needs no filename parsing:

```{r}
# example: merge Health survey with background variables
survey <- haven::read_sav("output/ch_merged.sav")

# read avars files; the period is the `wave` column in each file
bg_files <- list.files("data/avars/", pattern = "\\.sav$", full.names = TRUE)
bg_data  <- purrr::map_dfr(bg_files, function(f) {
  haven::read_sav(f) |> dplyr::mutate(fieldwork_ym = as.integer(wave))
})

merged <- dplyr::left_join(
  survey, bg_data,
  by = c("nomem_encr", "fieldwork_ym")
)
```
