---
title: "Getting Started with lissr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with lissr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Prerequisites

You need a registered account at the
[LISS Data Archive](https://www.dataarchive.lissdata.nl/).

## Store credentials

Store your username and password in the system keyring so lissr can
retrieve them automatically. The password is never saved in plain text or
in your R history.

```{r}
library(lissr)
liss_store_credentials("1234")
```

## Log in

```{r}
liss_login()
```

You will be prompted for a two-factor verification code sent to your email.

## Browse modules

```{r}
liss_modules()
liss_wave_matrix()
```

## Select and download

```{r}
selection <- liss_select()
liss_download(selection)
```

## Next steps

See [Merging LISS Panel Data](merge-workflow.html) for merging longitudinal data.
*(or run `vignette("merge-workflow", package = "lissr")` in the console)*
