---
title: "Longitudinal Panel Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Longitudinal Panel Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Why the merge engine matters for longitudinal work

The LISS panel surveys the same individuals annually. Exploiting that
panel structure — within-person change, fixed effects, event studies — is
what makes LISS valuable beyond a repeated cross-section. But it also
introduces complications that a cross-sectional workflow never encounters:

- **Prefix stripping**: wave ch07a stores self-rated health as `ch07a001`,
  while ch24q stores it as `ch24q001`. The merge engine strips the wave
  prefix so both become `s001`, making `bind_rows()` work.
- **Sentinel-code regime changes**: early waves use positive sentinels
  (999 = "don't know"), later waves switch to negative codes (-9). The
  recipe's harmonization rules recode both regimes to `NA` uniformly.
- **Instrument redesigns**: some items change wording or response scale
  between waves. The recipe's boundary rules flag these breaks with era
  indicators and comparability contracts — essential information for
  deciding whether to pool or split.
- **Variable renumbering**: questionnaire items sometimes get new suffix
  numbers. The recipe's crosswalk rules map old suffixes to new ones so
  you get a single harmonised column.

All of this is encoded in the YAML recipe and applied automatically.

## Step 1 — download all waves of a module

```{r download-all}
library(lissr)
library(dplyr)

liss_login()
bp <- liss_blueprint()

# all SPSS files for the Health module
health_files <- bp |>
  filter(module == "Health", type == "spss")

liss_download(health_files, .dir = "data/ch")
```

## Step 2 — merge all waves

```{r merge}
recipe <- liss_recipe("ch")
result <- merge_liss_module(recipe, data_dir = "data/ch", output_dir = "output")

panel <- result$data

# the stacked panel has one row per person-wave
panel |> count(wave_id) |> print(n = 20)
#> # A tibble: 18 × 2
#>    wave_id     n
#>    <chr>   <int>
#>  1 ch07a    6871
#>  2 ch08b    6386
#>  3 ch09c    6222
#>  ...  (18 waves, ch07a through ch25r; counts are illustrative)
```

## Step 3 — understand the panel structure

### Participation patterns

```{r participation}
# how many waves did each respondent participate in?
participation <- panel |>
  group_by(nomem_encr) |>
  summarise(n_waves = n_distinct(wave_id), .groups = "drop")

# distribution of participation (illustrative counts)
table(participation$n_waves)
#>    1    2    3   ...   16   17   18
#> 1842  801  602   ...  482  723 2103

# balanced sub-panel: respondents present in all 18 waves
balanced_ids <- participation |>
  filter(n_waves == max(n_waves)) |>
  pull(nomem_encr)

length(balanced_ids)
#> [1] 2103
```

### Attrition diagnostics

A common concern in panel studies is whether attrition is non-random. If
sicker people drop out, your estimated health trend is biased upward.

```{r attrition}
# tag respondents who appear in wave 1 but not in the final wave
first_wave <- "ch07a"
last_wave  <- "ch25r"

baseline <- panel |>
  filter(wave_id == first_wave) |>
  mutate(
    survived = nomem_encr %in%
      (panel |> filter(wave_id == last_wave) |> pull(nomem_encr))
  )

# compare baseline self-rated health between survivors and attriters
# (age is not a survey-module column; attach it from the Background
# Variables file if you need it in this comparison)
baseline |>
  group_by(survived) |>
  summarise(
    n         = n(),
    mean_srh  = mean(s001, na.rm = TRUE),
    .groups   = "drop"
  )
#> # A tibble: 2 × 3
#>   survived     n mean_srh
#>   <lgl>    <int>    <dbl>
#> 1 FALSE     4480     3.08
#> 2 TRUE      2103     3.25
```

If baseline health is significantly different between groups, consider
inverse-probability weighting or Heckman selection models.

## Step 4 — fixed-effects estimation

Person fixed effects absorb all time-invariant confounders (genetics,
childhood environment, stable personality). This is the workhorse model
for panel data.

```{r fixed-effects}
library(fixest)

# self-rated health (s001) regressed on a time-varying predictor with
# person and year fixed effects. pick the predictor from your module's
# codebook and use its harmonised suffix; `x_var` is a placeholder
panel <- panel |> mutate(x_var = s004)  # replace s004 with your item

fe_model <- fixest::feols(
  s001 ~ x_var | nomem_encr + wave_year,
  data = panel
)

summary(fe_model)
```

### Handling boundary breaks

The Health recipe declares boundary rules where the instrument changed:
for example, the e-cigarette and dental-hygienist items (suffixes
265-267) only exist from ch15h onward, and item 020 was reworded at
ch24q. Each rule materialises a flag column; you should respect them.

```{r boundary-check}
# inspect boundary flags created by the merge engine
flag_cols <- grep("_flag$|_period$|_era$|_present$", names(panel),
                  value = TRUE)
flag_cols
#> [1] "cancer_definition_era"        "alzheimer_definition_era"
#> [3] "ch_hca_items_present"         "ch_ecig_dental_items_present"
#> [5] "ch_years_not_working_present" "ch_covid_n2o_items_present"
#> [7] "ch_menopause_module_present"  "s020_wording_period"

# if your analysis touches the e-cigarette items, restrict to waves
# where the block was fielded — or include the flag as a control
panel |>
  filter(ch_ecig_dental_items_present) |>
  count(wave_year)
```

The merge report (`output/ch_merge_report.txt`) lists every comparability
contract and its recommended pooling strategy. Check it before deciding
to pool across eras.

## Step 5 — event-study / difference-in-differences

Panel data combined with a dated treatment allows event-study designs.
For example, suppose a policy change affected a subset of respondents
in 2016. You can estimate dynamic treatment effects:

```{r event-study}
# define treatment: respondents exposed to the policy, identified from
# a baseline covariate you attach (for example province of residence
# from the Background Variables file). `treated_ids` is your design's
# treatment group
treated_ids <- c( )  # fill from your design

panel <- panel |>
  mutate(
    treated  = as.integer(nomem_encr %in% treated_ids),
    post     = as.integer(wave_year >= 2016),
    rel_year = wave_year - 2016
  )

# event-study with staggered treatment
es_model <- fixest::feols(
  s001 ~ i(rel_year, treated, ref = -1) | nomem_encr + wave_year,
  data = panel
)

fixest::iplot(es_model, main = "Event study: self-rated health")
```

## Step 6 — growth curves and multilevel models

If you prefer a random-effects framework (e.g. for prediction or when
between-person variation is of interest), the stacked panel feeds
directly into `lme4`:

```{r growth-curve}
library(lme4)

# linear growth curve: health trajectories over time
panel <- panel |>
  mutate(year_centered = wave_year - 2007)

growth <- lme4::lmer(
  s001 ~ year_centered + (1 + year_centered | nomem_encr),
  data = panel
)

summary(growth)
```

## Working with a wave subset

Sometimes you only need a window of waves — for instance, the years
around a policy change. Rather than merging all 18 waves, trim the
recipe before merging:

```{r subset-waves}
recipe <- liss_recipe("ch")

# keep only 2015-2018 (ch skipped 2014; there is no ch14 wave)
target_waves <- c("ch15h", "ch16i", "ch17j", "ch18k")
recipe$wave_index <- purrr::keep(
  recipe$wave_index,
  ~ .x$id %in% target_waves
)

result <- merge_liss_module(recipe, data_dir = "data/ch", output_dir = "output")
```

This is faster and produces a smaller output file, but note that boundary
rules spanning the excluded waves will not fire.

## Audit trail

Every merge produces a JSONL log (`ch_merge_log.jsonl`) recording each
rule application with timestamps, row counts, and NA deltas. This gives
you a transparent audit trail for the Supplementary Materials section
of a paper — you can show exactly what transformations were applied to
the raw data.

```{r audit}
log <- jsonlite::stream_in(file("output/ch_merge_log.jsonl"), verbose = FALSE)

# how many values were recoded to NA?
log |>
  filter(action == "recode_to_na") |>
  summarise(total_recoded = sum(values_changed, na.rm = TRUE))
```

## Checklist for longitudinal LISS analyses

- [ ] Used `merge_liss_module()` or equivalent to stack waves with
  harmonised column names.
- [ ] Checked the merge report for comparability contracts and boundary
  flags before pooling across eras.
- [ ] Reported attrition rates and tested for selective dropout.
- [ ] Specified the estimator (FE, RE, FD) and justified it for your
  research question.
- [ ] If using person fixed effects, verified the outcome has sufficient
  within-person variation across waves.
- [ ] Attached time-varying demographics from the correct fieldwork-month
  background variables files.
- [ ] Cited the LISS wave identifiers and recipe version in the methods
  section.
