---
title: "Merging LISS Panel Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Merging LISS Panel Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

lissr includes a recipe-driven merge engine that processes YAML
specifications conforming to the Canonical Schema v1.1.0 (a strictly
additive extension of v1.0.0; recipes may declare either). Each recipe
encodes all merge-relevant decisions for a module: file patterns,
variable harmonization, boundary handling, comparability contracts, and
validation checks.

All merged output is written in SPSS `.sav` format to preserve variable
labels, value labels, and user-defined missing values.

## Merge downloaded modules

```{r}
library(lissr)

results <- merge_liss_modules(
  data_dir   = "liss",
  output_dir = "./output"
)
```

No module list or recipe paths are required. The batch detects supported
core-module wave files (`.sav`, `.zsav`, `.dta`, `.csv`), then loads only
the corresponding built-in recipes in alphabetical module-code order.
For each module, it checks its module-code subdirectory (for example
`liss/ch/`) if that directory exists, otherwise `data_dir` itself.
An existing empty module subdirectory therefore takes precedence over
files for that module in the parent folder. Filenames must retain their
module-wave prefix, such as `ch07a_`; existing recipe wave coverage still
determines which waves can be processed.
Modules without local files are omitted from the automatic batch.
Recipe, comparability, and data warnings from processed modules remain
visible. Finding a file does not establish that its format or contents are
analytically valid; review each module's report.

Discovery uses everything available in that folder, including older
downloads. It does not restrict the merge to the most recent `selection`
or infer which variables your analysis needs. Use a dedicated data folder
to keep a particular download separate.

Each module writes its own outputs. For example, Health produces:

- `ch_merged.sav` — merged data (SPSS format, preserving all labels)
- `ch_merge_log.jsonl` — audit-grade structured log
- `ch_merge_summary.json` — per-run summary (if enabled in recipe)
- `ch_merge_report.txt` — human-readable report

The returned list is named by module code, so `results$ch` contains the
Health result when local Health files are available. Each module result
additionally carries `provenance` (package, recipe,
and schema versions, per-input md5 hashes, release-selection decisions)
and a `valid_for_analysis` verdict, both reproduced in the report.

## Optional recipe controls

To merge only one module, or supply an already loaded custom recipe,
use the single-module helper. This is an alternative to the automatic
batch, not a preliminary step:

```{r}
recipe <- liss_recipe("ch")
result <- merge_liss_module(
  recipe,
  data_dir   = "liss/ch",
  output_dir = "./output"
)
```

The batch also accepts explicit recipe paths to restrict which recipes
run or to use custom recipes:

```{r}
custom_results <- merge_liss_modules(
  recipe_paths = "my_ch_recipe.yml",
  data_dir   = "liss",
  output_dir = "./output"
)
```

## Cross-module panel merge

After merging individual modules, combine them into one wide dataset.
Columns are prefixed with the module code to avoid collisions
(e.g. `ch_s004`, `cv_s004`).

```{r}
panel <- merge_liss_panel(results, write_to = "./output/liss_panel.sav")

# only respondent-years present in all modules
panel_inner <- merge_liss_panel(results, join_type = "inner")
```

## Validate recipes

```{r}
recipe <- liss_recipe("ch")
validate_recipe(recipe, "ch_merge_recipe.yml")
```

## Onboard a new wave

When a new wave is released, the onboarding helper automates most of the
checklist: variable diffs, candidate wave_index entry, expected-presence
checks, and boundary alerts.

```{r}
onboard_new_wave(
  recipe_path  = system.file("recipes", "ch_merge_recipe.yml", package = "lissr"),
  new_file     = "ch25r_EN_1.0p.sav",
  prev_wave_id = "ch24q"
)
```

## Background variables

The Background Variables file (`avars`) is a separate monthly snapshot,
not one of the ten core modules. Select the background month for the
survey items being analysed. Use `nomem_encr` as the respondent identifier,
never `nohouse_encr`: household assignments can change over time. When
stacking monthly snapshots, join on both `nomem_encr` and the selected
`fieldwork_ym` (YYYYMM), rather than on the respondent or calendar year
alone. Reject duplicate background keys and never match missing keys.

The source of `fieldwork_ym` is recipe-specific. The Politics and Values
(`cv`) recipe uses `_m` in eight single-part waves; in `cv16h`, group 0
uses `maandnr` and groups 1/2 use `maandnr_lang`; nine three-part waves
use `_m1` only. A missing designated date remains `NA`, without year
inference or fallback to another source. For three-part CV waves this is
part-1 timing, not an automatically appropriate background month for
part-2, part-3 or particular survey items. Check the relevant item or part
timing before choosing a monthly snapshot; the caveat must not depend on
a column label surviving a multi-wave merge.

The example below remains a Health join. Use `liss_blueprint(refresh = TRUE)` or `liss_select()` to select monthly
Background Variables ZIP releases (`module_id = 322`, `type = "archive"`,
`wave = YYYYMM`). Choose the required months and inspect the published
release descriptions and extracted contents. The catalogue does not verify
their data format or observed periods. This join requires locally available
SPSS files; acquire those separately if the selected ZIP contains another
format. The paths are illustrative. Each monthly SPSS file
records its period in `wave`; inspect the observed survey months and verify
the background periods before joining, rather than inferring them from
filenames:

```{r}
# example: merge Health survey with background variables
survey <- haven::read_sav("output/ch_merged.sav")
table(survey$fieldwork_ym, useNA = "ifany")

# read local monthly SPSS files; verify each observed period
bg_files <- list.files("data/avars/", pattern = "\\.sav$", full.names = TRUE)
stopifnot(length(bg_files) > 0L)
bg_data <- purrr::map_dfr(bg_files, function(f) {
  bg <- haven::read_sav(f)
  bg_month <- unique(as.integer(bg$wave))
  stopifnot(length(bg_month) == 1L, !anyNA(bg_month),
            bg_month %in% survey$fieldwork_ym)
  dplyr::mutate(bg, fieldwork_ym = as.integer(wave))
})

bg_keys <- bg_data[c("nomem_encr", "fieldwork_ym")]
stopifnot(!anyNA(bg_keys), !anyDuplicated(bg_keys))
merged <- dplyr::left_join(
  survey, bg_data,
  by = c("nomem_encr", "fieldwork_ym"), na_matches = "never"
)
stopifnot(nrow(merged) == nrow(survey))
```
